Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-03-26 doc/network.tex
should have commited earlier

Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution.
When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{champredon2015intrinsic}.

In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, measurements are ``censored'': we don't know what happens after the observation time.
These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions early in the epidemic.

Observed generation intervals also vary across space.
In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic. 
This means that contacts made late in infection are more likely to be ``masked'' due to infectious contacts that were made earlier.
Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

In this study, we explore spatiotemporal variation in observed generation-interval distributions.
We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
We discuss statistical approaches to estimating the effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
\swp{Come back and fix this}

@@

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
The population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
\jd{This is assuming that we know how to average across infected individuals. Do we need to assume that individual kernel properties are independent of risk of infection? If we have high-degree individuals, does it already break that assumption?}

There are two components to infection kernels: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
The basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
\begin{equation}
\RR_0 = \int K(t).
\end{equation}
\jd{OK, this one definitely looks problematic; that's not what happens with high degree.}
The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
\begin{equation}
g(\tau) = \frac{K(t)}{\RR_0}.
\end{equation}

\jd{Rewrite these equations a bit more consistently. What are we using for: index infection, current infection? $\tau$ is difference between them. Is it simpler just to use ' for dummies?}
In the homogeneous model, we assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
\begin{equation}
i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds.
\end{equation}
This model, also referred to as the renewal equation, can represent a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
Over a period of time where $S$ remains roughly constant, we would expect approximately exponential growth in $i$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
\begin{equation}
\frac{1}{\RR} = \int g(\tau) \exp(-r \tau) d\tau.
\end{equation}

@@

Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
Several studies have associated the backward generation-interval distribution with the early-epidemic censored generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}. 
The censored generation-interval distribution in fact averages over backward generation-interval distributions up until the observation time.
While we mathematically expect these underlying distributions to be the same (and thus the same distribution as their average) during the exponential growth phase, these assumptions can break down (due to susceptible depletion or other temporal change).
This, it is useful to conceptually distinguish the censored distribution the backward distribution.

The density of infections occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
\begin{equation}
i_{s-\tau}(s) = \RR i(s-\tau) g(\tau) S(s)
\end{equation}
The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
\jd{Is it nicer to keep $s$ in the numerator? Otherwise, why do we bother with $s$ at all?}
\begin{equation}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-x) g(x) d\tau}
\end{equation}

\jd{Can we just skip this \P, or put in an appendix or something?}
On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$ and is proportional to $\int_0^t i_{s-\tau}(s) ds$.
Then, the censored generation-interval distribution is given by
\begin{equation}\label{eq:obsg}
c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(\tau) \int_0^t i(s-\tau) S(s) ds d\tau}
\end{equation}
Thus, the censored generation-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
\begin{equation}
c_t(\tau) \propto \int_0^t b_s(\tau) S(s) ds.
\end{equation}
Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$.

@@

Generation intervals can be measured either forward (by tracing the infectees of an infected individual) or backward (by tracing the infector of an infected individual).
Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart.
Instead, we consider a slightly different scenario in which generation intervals are measured over multiple generations.
This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
In this case, the observed generation intervals are derived all infections that occured before the last date on which contact tracing was performed and are subject to right-censoring -- it is not possible to trace infections that did not happen yet.

When the epidemic is growing exponentially, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution.

@@

\subsection{Derivation of the individual-based method}

\pagebreak

\section{Junk}

First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (\fref{example}).
During the initial exponential growth period (approximately between 9 and 20 weeks), both population-level and individual-level methods yield similar estimates of mean generation interval and reproductive number.
However, we observe opposite trends after the growth period (approximately after 20 weeks). \jd{I kind of feel we're gettting off track here. The main focus should be on early fits, and we should be describing how cool \fref{cmp} is before we wander off into these woods.}
As the population-level method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
On the other hand, as the individual-level method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible is translated to underestimation of both mean generation interval and reproductive number.
we observe gradual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
Around the 14th week, reproductive number estimate based on the observed generation interval approximately matches the true reproductive number; this is due to initial overestimation of the initial growth rate and we generally expect reproductive to be underestimated. \swp{Maybe another figure in the supp}

\begin{figure}
\includegraphics[width=\textwidth]{../fig/example.pdf}
\caption{\swp{Need to re-do this figure; initial overestimation of little $r$ is weird}}
\label{fig:example}
\end{figure}

We compare accuracy and reliability of the two methods during the initial growth period using 100 stochastic simulations on a homogeneous network (Figure~\ref{fig:test}).
We find that the individual-level method provides the most consistent (least variable) estimates.
The population method overestimates reproductive number initially but this is likely to be driven by the initial overestimation of the growth rate as discussed above (Figure~\ref{fig:example}).

\begin{figure}
\includegraphics[width=\textwidth]{../fig/compare_methods.pdf}
\caption{\swp{Need to re-do this figure; initial overestimation of little $r$ is weird}.}
\label{fig:test}
\end{figure}

We also compare coverage probabilities of these methods. 
Coverage probabilities are defined as the proportion of confidence intervals that contain the true value in a repeated simulations; for example, a 95\% confidence interval should attain 95\% coverage probability by its definition.
Surprisingly, both methods only attain approximately 70\% coverage.
Even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution, making a wrong distributional assumption leads to narrow confidence intervals.
To confirm that the undercoverage is caused by distribuional assumption, we simulated an epidemic using gamma intrinsic generation-interval and found that fitting the true shape yields nominal coverage (see Appendix).
These results are particularly alarming because shape of the generation-interval distribution have often been assumed in outbreak analyses and it is impossible to know the true shape of the distribution [CITE].

\clearpage

The first method is a non-parametric method.
Recall that the observed generation-interval distribution is a weighted intrinsic generation-interval distribution (equation~\ref{eq:obsg}). 
Then, the intrinsic generation-interval distribution can be obtained by taking the inverse weight:
\begin{equation}
g(\tau) \propto c_t(\tau) \frac{1}{\int_{0}^t i(s-\tau) S(s) ds}
\end{equation}
However, this method requires a knowledge of susceptible dynamics and may difficult to use in practice.
Instead, during the exponential period, the generation-interval distribution is
\begin{equation}
g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau)
\end{equation}
and the reproductive number is
\begin{equation}
\RR = \int \tsub{c}{exp}(\tau) \exp(r\tau) d\tau.
\end{equation}
Note that applying the Lotka-Euler equation directly to the observed generation intervals will always underestimate the reproductive number:
\begin{equation}
\int_0^\infty \tsub{g}{exp}(\tau) \exp(r\tau) d\tau > \left(\int_0^\infty \tsub{g}{exp}(\tau). \exp(-r\tau) d\tau\right)^{-1}
\end{equation}

Finally, we obtain an estimator for mean generation interval and the reproductive number:
\begin{equation}
\begin{aligned}
\hat{G} &= \frac{\sum_{i} \exp(r c_i) c_i}{\sum_{i} \exp(r c_i)},\\
\hat{\RR} &= \sum_{i} \exp(r c_i),
\end{aligned}
\end{equation}
where $r$ is the exponential growth rate and $c_i$ is an individual generation interval sample.

\section{Appendix}

These are notes for now ... Need to move them elsewhere and change notation... but this is part of our figure for empirical network ...

$$
1 = \kappa \phi_L(\alpha) \frac{\lambda^{(net)}}{\alpha + \lambda^{(net)}} (1 - \phi_I(\alpha+\lambda^{(net)}))
$$
where in our specific example,
$$
\phi_L(\alpha) = \frac{4 \delta^2}{(2 \delta + \alpha)^2}
$$
and so we have
$$
1 = \frac{4 \delta^2}{(2 \delta + \alpha)^2} \frac{\kappa \lambda^{(net)}}{\lambda^{(net)} + \alpha + \gamma},
$$
which suggests that
$$
\lambda^{(net)} = \frac{(2\delta + \alpha)^2 (\alpha + \gamma)}{(\kappa - 1) 4 \delta^2 - 4 \delta \alpha - \alpha^2}
$$
We can substitute into the following equation to obtain the $r$ and $\RR$ relationship:
$$
\RR = \kappa \frac{\lambda^{(net)}}{\lambda^{(net)} + \gamma}
$$

>>>>

\jd{Consider rewriting to use early-late as the first-level structure. Early is the part that people focus on most. We could even throw in censorship.}
\swp{Not sure what you mean... I tried to focus on early-late throughout the manuscript}
Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution.
When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{champredon2015intrinsic}.

In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, these measurements are \emph{censored} (or right-censored) at the observation time.
These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions early in the epidemic.

Observed generation intervals also vary across space.
In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic. 
This means that contacts made late in infection are more likely to be ``masked'' due to infecious contacts that were made earlier.
Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

In this study, we explore the spatiotemporal variation in the observed generation-interval distributions obtained from contact tracing.
We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
We discuss statistical approaches to estimating the effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
\swp{Come back and fix this}

@@

Generation-interval distributions are often viewed from a population-level pespective but we can distinguish population-level distributions from individual-level distributions (\fref{indpop}); 
making this distinction clear will be particularly useful when we discuss spatial components.
An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}.
The population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
\jd{Do we need to worry about assumptions for the rest of this \P? Once we have an aspect space we need to assume that individual properties are independent of risk of infection. Also, since that's wrong on our networks, do we have to figure out why we're approximately saved? I guess because of scaling?}
\swp{What do you mean by individual properties or even scaling?}

There are two components to infection kernels: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
The basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
\begin{equation}
\RR_0 = \int K(t).
\end{equation}
The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
\begin{equation}
g(\tau) = \frac{K(t)}{\RR_0}.
\end{equation}

We assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
\begin{equation}
i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds.
\end{equation}
This model, also referred to as the renewal equation, can represent a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
Over a period of time where $S$ remains roughly constant, we would expect approximately exponential growth in $i$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
\begin{equation}
\frac{1}{\RR} = \int g(\tau) \exp(-r \tau) d\tau.
\end{equation}

@@

\jd{Mathematically, early censored and early backward intervals are the same, but conceptually they're just different. In particular, censoring occurs even with forward tracing.}
\jd{During exponential growth, the theoretical distributions are the same; maybe we should do a tiny bit of that math here so that we can talk more clearly about it.}
\jd{We need to think about exposition. Maybe simple formulae for all of our interval types, both in general and under the assumption of exponential growth.}
\jd{We should be saying that backward $\to$ censored in the limit at the same time we talk about the association below.}
\swp{I tried}

Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}, but the backward generation-interval distribution takes into account infections that are only a single generation apart from the focal time.
Instead, contact tracing can be performed over multiple generations, tracing all the way back to the primary case.
The backward generation-interval distribution is mathematically equivalent to the censored generation-interval distribution only during the exponential growth phase, when depletion of the susceptible population is negligible.
However, the exponential growth assumption breaks down rapidly, and it is important to be able to conceptually distinguish the censored generation-interval from the backward generation-interval.
\jd{I hope that none of these papers actually make the mistake that you imply, it's all about the limit.}
\swp{Sure, in the limit they're all same. But we're often not in the limit and contact tracing can be performed beyond the limit. We can make this point.}

The number of infection occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
\begin{equation}
i_{s-\tau}(s) = \RR i(s-\tau) g(\tau) S(s)
\end{equation}
The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
\begin{equation}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-x) g(x) d\tau}
\end{equation}
On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$ and is proportional to $\int_0^t i_{s-\tau}(s) ds$.
Then, the censored generation-interval distribution is given by
\begin{equation}\label{eq:obsg}
c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(\tau) \int_0^t i(s-\tau) S(s) ds d\tau}
\end{equation}
Alternatively, the censored generatio-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
\begin{equation}
c_t(\tau) \propto \int_0^t b_s(\tau) S(s) ds.
\end{equation}
Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$.

@@

Generation intervals can be measured either forward (by tracing the infectees of an infected individual) or backward (by tracing the infector of an infected individual).
Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart.
Instead, we consider a slightly different scenario in which generation intervals are measured over multiple generations.
This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
In this case, the observed generation intervals are derived all infections that occured before the last date on which contact tracing was performed and are subject to right-censoring -- it is not possible to trace individuals that have not been incted yet.

When the epidemic is growing exponentially, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-censoring.
Near the end of an epidemic, the effect of right-censoring becomes negligible but the observed generation intervals are still shorter on average than the intrinsic generation intervals.

We further classify susceptible depletion into three levels: egocentric, local, and global.
Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
Accounting for the egocentric effect allows us to reinterpret the results by Trapman et al using arguments based on generation intervals.
Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
Global susceptible depletion refers to the overall decrease in the susceptible pool in the population.

Susecptible depletion happening at all three levels shortens realized generation intervals but acts on different time scales.

Limitations
\begin{itemize}
	\item Exponential growth; polynomial
	\item Extending the framework (accounting for control)
	\item Low coverage of gamma
	\item Serial interval
\end{itemize}

@@
